SCPCL000822 - Validation of tumor cells

### 2024-06-14 analyses/cell-type-ewings/cell_type_annotations/SCPCS000490/SCPCL000822_tumor-cell-validation.Rmd
rename unknown to ambiguous and no extra classification

@@

```{r}
# create a combined classification between inferCNV and copyKAT
# cells that are classified as tumor by both are kept as tumor
# classified by both as normal are normal 
# in between are classified as unknown
classification_df <- classification_df |> 
  dplyr::mutate(
    combined_cnv_classification = dplyr::case_when(
      # only tumor if both call tumor 
      cnv_proportion_classification == "Tumor" & copykat_classification == "Tumor" ~ "Tumor",
      # only normal if both call normal and/or reference
      cnv_proportion_classification == "Normal" & copykat_classification == "Normal" ~ "Normal",
      cnv_proportion_classification == "Reference" & copykat_classification == "Normal" ~ "Normal",
      # if normal/tumor mismatch set to unknown 
      .default = "Unknown"
    ))
```

@@

annotation <- ComplexHeatmap::columnAnnotation(
  classification = annotation_df$combined_cnv_classification,
  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))
```

@@

```{r}
# need to make a new annotation, since we want to annotate the rows here 
annotation <- ComplexHeatmap::rowAnnotation(classification = annotation_df$combined_cnv_classification,
                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))

@@

It looks like we have a pretty clear distinction between the `proportion_scaled_dupli_chr8` for normal and tumor cells and we could call any cell with a value > 0.25 as tumor cells and those that are less than 0.25 as normal cells.

```{r}
new_classification_df <- classification_df |> 
  # remove duplicat
  dplyr::select(-c(gene_symbol, ensembl_gene_id, gene_expression)) |> 
  dplyr::distinct() |> 
  dplyr::mutate(
    combined_cnv_classification = dplyr::case_when(
      combined_cnv_classification == "Unknown" & proportion_scaled_dupli_chr8 >= 0.25 ~ "Tumor",
      combined_cnv_classification == "Unknown" & proportion_scaled_dupli_chr8 < 0.25 ~ "Normal",
      .default = combined_cnv_classification
    )
  )
```

```{r}
# account for new annotations 
annotation <- ComplexHeatmap::rowAnnotation(classification = new_classification_df$combined_cnv_classification,
                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Unknown" = "grey")))

# recreate the cnv heatmap 
plot_cnv_heatmap(new_classification_df,
                 cnv_col = "proportion_scaled_dupli",
                 annotation = annotation,
                 legend_title = "Copy number \ngain")
```

It looks like we now have a prety good separation between tumor and normal cells based mostly on Chr8 gain. 
Before we finish, we'll make a UMAP showing where the final annotations are.

```{r}
# look at which cells are tumor/normal
new_classification_df |> 
  ggplot(aes(x = UMAP1, y = UMAP2, color = combined_cnv_classification)) +
  geom_point(alpha = 0.5, size = 0.5)
```

>>>>

### Process input files

@@

```{r}
# create a combined classification between inferCNV and copyKAT
# cells that are classified as tumor by both are kept as tumor
# classified by both as normal are normal 
# in between are classified as unknown
classification_df <- classification_df |> 
  dplyr::mutate(
    combined_cnv_classification = dplyr::case_when(
      # only tumor if both call tumor 
      cnv_proportion_classification == "Tumor" & copykat_classification == "Tumor" ~ "Tumor",
      # only normal if both call normal and/or reference
      cnv_proportion_classification == "Normal" & copykat_classification == "Normal" ~ "Normal",
      cnv_proportion_classification == "Reference" & copykat_classification == "Normal" ~ "Normal",
      # if normal/tumor mismatch set to unknown 
      .default = "Ambiguous"
    ), 
    # set factor order to use for plotting
    combined_cnv_classification = forcats::fct_relevel(combined_cnv_classification, "Tumor", "Normal", "Ambiguous"))
```

@@

annotation <- ComplexHeatmap::columnAnnotation(
  classification = annotation_df$combined_cnv_classification,
  col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))
```

@@

```{r}
# need to make a new annotation, since we want to annotate the rows here 
annotation <- ComplexHeatmap::rowAnnotation(classification = annotation_df$combined_cnv_classification,
                                            col = list(classification = c("Tumor" = "#00274C", "Normal" = "#FFCB05", "Ambiguous" = "grey")))

@@

It looks like we have a pretty clear distinction between the `proportion_scaled_dupli_chr8` for normal and tumor cells, but there are a few bumps indicating some normal cells do have higher values for CNVs at Chr8 that are more similar tumor cells. 
Because of this, we will keep the "Ambiguous" cells as is and not try to classify them.
